Different forms of uncertainty differently affect the evolution of social learning

### 2022-08-25 ms.tex
updated convergence in ms, updated expected payoffs component figs
updated figure 5->3 and did supplement S tables and figs

First, we confirmed that social learning evolves when the expected payoffs to social
learners exceed those of asocial learners, which we write $\meansoc > \meanasoc$.
In the case of $L=1$ we can calculate $\meanasoc$ directly: it is just the expected
payoff across all behaviors. For all other cases, we calculate expected social and asocial payoffs by respectively setting
$\sum_i s_i = N$ or $\sum_i s_i = 0$ at model initialization. These stay constant since we do not allow for mutation. These are the payoffs to a homogeneous population of either all social learners or all asocial learners.  Figure~\ref{fig:payoffs} shows these expected payoffs, normalized by effective lifespan $L$), as well as the observed average payoffs in our simulations. \ps{Is this right? I'm having trouble understanding what the solid lines in Figs 3 and 5 are. I'm also debating the wisdom of separating these figures. Maybe we put the 3x3 back in for Fig 3?}
In general, we observed that social learning evolved in our simulations when the expected payoffs exceeded expected individual-learner
payoffs (Figure~\ref{fig:payoffs}). 
However, we also found that populations of initially all social learners performed worse
in many uncertainty contexts than either the actual payoffs accumulated by 
model agents, and lower than even the expected individual-level payoffs,
even though social learning evolved in many of the same contexts in our model,
especially when $\pilow = 0.8$ 
(Figure~\ref{fig:fullMeanPrevNetPayoffs}, bottom row). \ps{What is the point of this last finding? I'm having trouble figuring out what all this has to do with drift.}

\begin{figure}
  \caption{Expected payoff rates for homogeneous populations of social or asocial learners. Dashed lines with no markers are expected payoff rates for a population of individual learners. Dashed lines with diamond markers are expected payoff rates to a population of social learners. Solid lines with circle markers are the actual average payoff rates for agents in our evolutionary simulations.  
%  Comparison of expected payoffs from our computational anaysis (solid     lines with circles), expected individual payoffs with no social learning     (dashes without markers), and expected payoffs for all-social learner     populations. 
Only four plots are shown to enhance detail; see supplement for
    five additional plots.
} 
  \label{fig:payoffs}
\centering
    \includegraphics[width=0.85\textwidth]{Figures/meanNetPayoffs.pdf}
\end{figure}

@@

Our model necessarily made simplifying assumptions. These choices were judiciously made to address the main question of how uncertainty affects the evolution of social learning in general; however there exist alternative empirically-valid choices we could have made, the future exploration of which will be important to the development of a more nuanced theory of social learning.  % that may not always hold in more specific situations, though they are empirically justified to be general across taxa. 
For example, we considered only success-biased
learning, although conformity or other context biases may sometimes be the social learning mechanism~\cite{BoydRicherson1985,Muthukrishna2016a,Smaldino2018b}.  Conformity may lead to more drift
overall, since success-biased teacher selection, which we assumed, maximizes the potential benefit to social learners. We also assumed that the number of behaviors and
their payoffs were constant within a given simulation, but this fails to account for evolutionary feedback which
creates new behavioral opportunities as time progresses, e.g.\ via niche
construction~\cite{Smaldino2012a,Heras-Escribano2020} or cumulative cultural evolution
supported by dynamic network structure~\cite{Smolla2019,Derex2020}.  Group structure
and processes such as homophily and in-group bias/out-group aversion could inhibit the
evolution of social learning since successful outgroup teachers could be cut off from
in-group learners~\cite{Jackson2012}. We assumed agents choose behaviors via softmax
search, but this is an underpowered algorithm compared to the real human individual learning~\cite{Schulz2020a,Wu2022}.  More powerful individual learning could further
support social learning.  These considerations reveal new opportunities to enhance theoretical nuance by adjusting or extending the present model for other contexts.

@@

\appendix

\section{Additional main figure: Full net payoff comparison for all nine uncertainty conditions} 
\ps{This should be part of the supplement. Also, all supplementary figures and tables should be preceded by ``S".}
In the main text we showed a $2\times2$ array of plots comparing the observed
mean net payoffs obtained by agents in our simulations, and in two auxiliary
sets of simulations that set all agents to either have the social learning
trait or not, which remained the case for the entire simulation since there
are no mutations in this model (Figure~\ref{fig:payoffs}). Here we show the full $3\times3$ array of
all $(\pilow, B)$ combinations shown for the main results. We see that
populations of all social learners especially underperform initially mixed
populations that evolve to be social learners 
(Figure~\ref{fig:fullMeanPrevNetPayoffs}, bottom row). This highlights the
importance of diversity for behavioral optimization, even if it is transient.

\begin{figure}
  \caption{Expected payoff rates for homogeneous populations of social or asocial learners. Dashed lines with no markers are expected payoff rates for a population of individual learners. Dashed lines with diamond markers are expected payoff rates to a population of social learners. Solid lines with circle markers are the actual average payoff rates for agents in our evolutionary simulations.  When $\pilow = 0.8$, all-social learner populations underperform
    since social learners are not subject to an initial evolutionary bottleneck
    due to the presence of asocial learners, and behavioral misinformation
    is allowed to circulate more freely than in the base case tested by 
    our model with an initially even distribution of social and asocial learners.
}
  \label{fig:fullMeanPrevNetPayoffs}
  \centering
    \includegraphics[width=\textwidth]{Figures/supplement/fullMeanPrevNetPayoffs.pdf}
\end{figure}

@@

\section{Supplemental supporting information}

\subsection{Convergence information}

\begin{table}[h] \caption{Maximum number of iterations = 20000.} \label{tab:convergence} \centering
  \begin{tabular}{cccc} 
    \toprule $B$ & \# not at fixation (out of 132k) & \% not fixated \\ 
    \midrule  2  & 192  & 0.1 \% \\ 
              4  & 274  & 0.2 \% \\ 
              10 & 2935 & 2.0 \% \\ 
    \bottomrule \end{tabular} 
\end{table}

\subsection{Population size sensitivty analysis}

@@

\begin{figure}
  \centering
  \addtocounter{figure}{-1}
  \caption{
	Sensitivity analysis of the main results for different population
	sizes, $N=50,200,1000$. Recall $N=100$ was used to generate main 
	text results.
  }
  \label{fig:populationSensitivity}
  \begin{subfigure}{\textwidth}
	\caption{$N=50$}
	\includegraphics[width=\textwidth]{Figures/supplement/nagents=50/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

@@

\subsection{Number of prospective teachers sensitivty analysis}

@@

\vspace{-3em}
\begin{figure}
  \centering
  \addtocounter{figure}{-1}
  \caption{Number of prospective teachers sensitivity analysis for $N_T=2,10,20$. Recall
  $N_T=5$ was used to generate main text results.}
  \label{fig:nteachersSensitivity}
  \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{$N_T = 2$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=2/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$N_T = 10$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=10/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$N_T = 20$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=20/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

@@

\subsection{Softmax parameter sensitivity analysis}

@@

\vspace{-3em} \begin{figure} %\addtocounter{figure}{-1} 
  \centering
  \caption{Sensitivity analysis of the main results for the softmax parameter $\beta = 100$ and
  \addtocounter{figure}{-1}
  $\beta=1$. Recall the main results were obtained with $\beta = 10$.}
  \label{fig:softmaxSensitivity} \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{$\beta = 1$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=1.0/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$\beta = 100$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=0.01/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

>>>>

First, we confirmed that social learning evolves when the expected payoffs to social
learners exceed those of asocial learners, which we write $\meansoc > \meanasoc$.
In the case of $L=1$ we can calculate $\meanasoc$ directly: it is just the expected
payoff across all behaviors. For all other cases, we calculate expected social and asocial payoffs by respectively setting
$\sum_i s_i = N$ or $\sum_i s_i = 0$ at model initialization. These stay constant since we do not allow for mutation. These are the payoffs to a homogeneous population of either all social learners or all asocial learners.  Figure~\ref{fig:payoffs} shows these expected payoffs, normalized by effective lifespan $L$), as well as the observed average payoffs in our simulations. \ps{Is this right? I'm having trouble understanding what the solid lines in Figs 3 and 5 are. I'm also debating the wisdom of separating these figures. Maybe we put the 3x3 back in for Fig 3?}
In general, we observed that social learning evolved in our simulations when the expected payoffs exceeded expected individual-learner
payoffs (Figure~\ref{fig:payoffs}). 
% However, we also found that populations of initially all social learners performed worse
% in many uncertainty contexts than either the actual payoffs accumulated by 
% model agents, and lower than even the expected individual-level payoffs,
% even though social learning evolved in many of the same contexts in our model,
% especially when $\pilow = 0.8$ 
% (Figure~\ref{fig:fullMeanPrevNetPayoffs}, bottom row). \ps{What is the point of this last finding? I'm having trouble figuring out what all this has to do with drift.}

% \begin{figure}
%   \caption{Expected payoff rates for homogeneous populations of social or asocial learners. Dashed lines with no markers are expected payoff rates for a population of individual learners. Dashed lines with diamond markers are expected payoff rates to a population of social learners. Solid lines with circle markers are the actual average payoff rates for agents in our evolutionary simulations.  
% %  Comparison of expected payoffs from our computational anaysis (solid     lines with circles), expected individual payoffs with no social learning     (dashes without markers), and expected payoffs for all-social learner     populations. 
% Only four plots are shown to enhance detail; see supplement for
%     five additional plots.
% } 
%   \label{fig:payoffs}
% \centering
%     \includegraphics[width=0.85\textwidth]{Figures/meanNetPayoffs.pdf}
% \end{figure}

\begin{figure}
  \caption{Expected payoff rates for homogeneous populations of social or asocial learners. Dashed lines with no markers are expected payoff rates for a population of individual learners. Dashed lines with diamond markers are expected payoff rates to a population of social learners. Solid lines with circle markers are the actual average payoff rates for agents in our evolutionary simulations.  When $\pilow = 0.8$, all-social learner populations underperform
    since social learners are not subject to an initial evolutionary bottleneck
    due to the presence of asocial learners, and behavioral misinformation
    is allowed to circulate more freely than in the base case tested by 
    our model with an initially even distribution of social and asocial learners.
    \mt{Correct or nah?}
}
  \label{fig:payoffs}
  \centering
    \includegraphics[width=\textwidth]{Figures/supplement/fullMeanPrevNetPayoffs.pdf}
\end{figure}

@@

Our model necessarily made simplifying assumptions. These choices were judiciously made to address the main question of how uncertainty affects the evolution of social learning in general; however there exist alternative empirically-valid choices we could have made, the future exploration of which will be important to the development of a more nuanced theory of social learning.  % that may not always hold in more specific situations, though they are empirically justified to be general across taxa. 
For example, we considered only success-biased
learning, although conformity or other context biases may sometimes be the social learning mechanism~\cite{BoydRicherson1985,Muthukrishna2016a,Smaldino2018b}.  Conformity may lead to more drift
overall, since success-biased teacher selection, which we assumed, maximizes the potential benefit to social learners. We also assumed that the number of behaviors and
their payoffs were constant within a given simulation, but this fails to account for evolutionary feedback which
creates new behavioral opportunities as time progresses, e.g.\ via niche
construction~\cite{Smaldino2012a,Heras-Escribano2020} or cumulative cultural evolution
supported by dynamic network structure~\cite{Smolla2019,Derex2020}.  Group structure
and processes such as homophily and in-group bias/out-group aversion could inhibit the
evolution of social learning since successful outgroup teachers could be cut off from
in-group learners~\cite{Golub2012}. We assumed agents choose behaviors via softmax
search, but this is an underpowered algorithm compared to the real human individual learning~\cite{Schulz2020a,Wu2022}.  More powerful individual learning could further
support social learning.  These considerations reveal new opportunities to enhance theoretical nuance by adjusting or extending the present model for other contexts.

@@

%%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
\pagebreak
\begin{center}
  \textbf{\Large \textsf{Supplement}}
\end{center}
%%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
\setcounter{equation}{1}
\setcounter{figure}{0}
\setcounter{section}{0}
\setcounter{table}{0}
\setcounter{page}{1}
\makeatletter
\renewcommand{\theequation}{S\arabic{equation}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\thepage}{S\arabic{page}}
% \renewcommand{\bibnumfmt}[1]{[S#1]}
% \renewcommand{\citenumfont}[1]{S#1}
%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%

% \section{Additional main figure: Full net payoff comparison for all nine uncertainty conditions} 
% \ps{This should be part of the supplement. Also, all supplementary figures and tables should be preceded by ``S".}
% In the main text we showed a $2\times2$ array of plots comparing the observed
% mean net payoffs obtained by agents in our simulations, and in two auxiliary
% sets of simulations that set all agents to either have the social learning
% trait or not, which remained the case for the entire simulation since there
% are no mutations in this model (Figure~\ref{fig:payoffs}). Here we show the full $3\times3$ array of
% all $(\pilow, B)$ combinations shown for the main results. We see that
% populations of all social learners especially underperform initially mixed
% populations that evolve to be social learners 
% (Figure~\ref{fig:fullMeanPrevNetPayoffs}, bottom row). This highlights the
% importance of diversity for behavioral optimization, even if it is transient.

% \newpage

\section{Convergence information}

\begin{table}[h] \caption{Maximum number of iterations = 20000.} \label{tab:convergence} \centering
  \begin{tabular}{cccc} 
    \toprule $B$ & \# not at fixation (out of 132k) & \% not fixated \\ 
    \midrule  2  & 0  & 0.0 \% \\ 
              4  & 0  & 0.0 \% \\ 
              10 & 20 & 0.02 \% \\ 
    \bottomrule \end{tabular} 
\end{table}

@@

\section{Population size sensitivty analysis}

@@

\begin{figure}
  \centering
  % \addtocounter{figure}{-1}
  \caption{
	Sensitivity analysis of the main results for different population
	sizes, $N=50,200,1000$. Recall $N=100$ was used to generate main 
	text results.
  }
  \label{fig:populationSensitivity}
  \begin{subfigure}{\textwidth}
	\caption{$N=50$}
	\includegraphics[width=\textwidth]{Figures/supplement/nagents=50/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

@@

\section{Number of prospective teachers sensitivty analysis}

@@

\vspace{-3em}
\begin{figure}
  \centering
  % \addtocounter{figure}{-1}
  \caption{Number of prospective teachers sensitivity analysis for $N_T=2,10,20$. Recall
  $N_T=5$ was used to generate main text results.}
  \label{fig:nteachersSensitivity}
  \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{$N_T = 2$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=2/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$N_T = 10$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=10/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$N_T = 20$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=20/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

@@

\section{Softmax parameter sensitivity analysis}

@@

\vspace{-3em} \begin{figure} %\addtocounter{figure}{-1} 
  \centering
  \caption{Sensitivity analysis of the main results for the softmax parameter $\beta = 100$ and
  % \addtocounter{figure}{-1}
  $\beta=1$. Recall the main results were obtained with $\beta = 10$.}
  \label{fig:softmaxSensitivity} \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{$\beta = 1$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=1.0/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$\beta = 100$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=0.01/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
